Validation of GMEA

### 2023-09-15 figs/step_01c_splithcy.Rmd
split+analyse

suppressPackageStartupMessages({
  library("missMethyl")
  library("kableExtra")
  library("mitch")
  library("eulerr")
  library("IlluminaHumanMethylation450kmanifest")
  library("IlluminaHumanMethylation450kanno.ilmn12.hg19")
  library("tictoc")
})

@@

@@

design <- readRDS("dm3_design.Rds")
dim(design)

dm <- readRDS("dm3_mx.Rds")
dim(dm)

@@

Split the data.

@@

dess <- design[c(gxs),]
dms <- dm[,c(gxs)]

desb <- design[c(gxb),]
dmb <- dm[,c(gxb)]

@@

@@

## GMEA

Now with GMEA.

Start with gathering the annotation set.

Histogram of limma t-statistics.

```{r,hist}

dm <- dm[,c("UCSC_RefGene_Name","t")]
hist(dm$t,breaks=seq(from=-6,to=6,by=1)) ; grid()

Does 1-sample t-test work better?

```{r,agg_tt}

agg2 <- function(dm) {
  gn <- unique(unlist(strsplit( dm$UCSC_RefGene_Name ,";")))
  gnl <- strsplit( dm$UCSC_RefGene_Name ,";")
  gnl <- mclapply(gnl,unique,mc.cores=CORES)
  dm$UCSC_RefGene_Name <- gnl
  l <- mclapply(1:nrow(dm), function(i) {
    a <- dm[i,]
    len <- length(a[[1]][[1]])
    tvals <- as.numeric(rep(a["t"],len))
    genes <- a[[1]][[1]]
    data.frame(genes,tvals)
  },mc.cores=CORES)
  df <- do.call(rbind,l)
  keep <- names(which(table(df$genes)>1))
  df <- df[df$genes %in% keep,]
  gn <- unique(df$genes)
  gme_res <- lapply( 1:length(gn), function(i) {
    g <- gn[i]
    tstats <- df[which(df$genes==g),"tvals"]
    myn <- length(tstats)
    mymean <- mean(tstats)
    mymedian <- median(tstats)
    tselfcont <- t.test(tstats)
    res <- c("gene"=g,"nprobes"=myn,"mean"=mymean,"median"=mymedian,
      "p-value(sc)"=tselfcont$p.value)
  } )
  gme_res_df <- do.call(rbind, gme_res)
  rownames(gme_res_df) <- gme_res_df[,1]
  gme_res_df <- gme_res_df[,-1]
  tmp <- apply(gme_res_df,2,as.numeric)
  rownames(tmp) <- rownames(gme_res_df)
  gme_res_df <- as.data.frame(tmp)
  gme_res_df$sig <- -log10(gme_res_df[,4])
  gme_res_df <- gme_res_df[order(-gme_res_df$sig),]
  gme_res_df$`fdr(sc)` <- p.adjust(gme_res_df$`p-value(sc)`)
  out <- list("df"=df,"gme_res_df"=gme_res_df)
  return(out)
}

tic()
dmagg2 <- agg2(dm)
toc()
head(dmagg2$gme_res_df) %>% kbl(caption="top gmea genes (t-test)") %>% kable_paper("hover", full_width = F)

hist(dmagg2$gme_res_df$median,breaks=30) ; grid()

Conduct a 1-sample t-test.

```{r,t_enrichment_test1}

ttenrich <- function(m,genesets) {
  res <- mclapply( 1:length(genesets), function(i) {
    gs <- genesets[i]
    name <- names(gs)
    n_members <- length(which(rownames(m) %in% gs[[1]]))
    if ( n_members > 4 ) {
      tstats <- m[which(rownames(m) %in% gs[[1]]),]
      myn <- length(tstats)
      mymean <- mean(tstats)
      mymedian <- median(tstats)
      wt <- t.test(tstats)
      res <- c(name,myn,mymean,mymedian,wt$p.value)
    }
  } , mc.cores = CORES/2)
  res_df <- do.call(rbind, res)
  rownames(res_df) <- res_df[,1]
  res_df <- res_df[,-1]
  colnames(res_df) <- c("n_genes","t_mean","t_median","p.value")
  tmp <- apply(res_df,2,as.numeric)
  rownames(tmp) <- rownames(res_df)
  res_df <- tmp
  res_df <- as.data.frame(res_df)
  res_df <- res_df[order(res_df$p.value),]
  res_df$logp <- -log10(res_df$p.value )
  res_df$fdr <- p.adjust(res_df$p.value,method="fdr")
  res_df[order(abs(res_df$p.value)),]
  return(res_df)
}

tic()
tres <- ttenrich(m=m,genesets=gs_symbols)
toc()

head(tres,30) %>%
  kbl(caption = "Top significant genesets") %>%
  kable_paper("hover", full_width = F)

head(tres[order(-abs(tres$t_median)),],30) %>%
  kbl(caption = "Top effect size genesets") %>%
  kable_paper("hover", full_width = F)

sig <- subset(tres,fdr<0.05)
nrow(sig)

head(sig[order(-abs(sig$t_median)),],30) %>%
  kbl(caption = "Top effect size genesets after FDR filtering") %>%
  kable_paper("hover", full_width = F)

plot(tres$t_median,-log10(tres$p.value),pch=19,col="gray")
grid()
points(sig$t_median,-log10(sig$p.value),pch=19,col="red")

tsig <- sig

@@

msig_up <- subset(msig,s.dist>0)$set
msig_dn <- subset(msig,s.dist<0)$set

tsig_up <- rownames(subset(tsig,t_median>0))
tsig_dn <- rownames(subset(tsig,t_median<0))

v1 <- list("mitch up"=msig_up, "mitch dn"=msig_dn,
  "ttest up"=tsig_up,"ttest dn"=tsig_dn)

@@

Split data set into 2 halves to show consistency.

>>>>

suppressPackageStartupMessages({
  library("mitch")
  library("limma")
  library("kableExtra")
  library("eulerr")
  library("IlluminaHumanMethylation450kmanifest")
  library("IlluminaHumanMethylation450kanno.ilmn12.hg19")
  library("tictoc")
})

@@

## Load gene sets

## Load annotation data

## Load functions for GMEA

```{r,gmeafunc}

agg <- function(dm,cores=1) {
  gn <- unique(unlist(strsplit( dm$UCSC_RefGene_Name ,";")))
  gnl <- strsplit( dm$UCSC_RefGene_Name ,";")
  gnl <- mclapply(gnl,unique,mc.cores=cores)
  dm$UCSC_RefGene_Name <- gnl
  l <- mclapply(1:nrow(dm), function(i) {
    a <- dm[i,]
    len <- length(a[[1]][[1]])
    tvals <- as.numeric(rep(a["t"],len))
    genes <- a[[1]][[1]]
    data.frame(genes,tvals)
  },mc.cores=cores)
  df <- do.call(rbind,l)
  keep <- names(which(table(df$genes)>1))
  df <- df[df$genes %in% keep,]
  gn <- unique(df$genes)
  gme_res <- lapply( 1:length(gn), function(i) {
    g <- gn[i]
    tstats <- df[which(df$genes==g),"tvals"]
    myn <- length(tstats)
    mymean <- mean(tstats)
    mymedian <- median(tstats)
    tselfcont <- t.test(tstats)
    res <- c("gene"=g,"nprobes"=myn,"mean"=mymean,"median"=mymedian,
      "p-value(sc)"=tselfcont$p.value)
  } )
  gme_res_df <- do.call(rbind, gme_res)
  rownames(gme_res_df) <- gme_res_df[,1]
  gme_res_df <- gme_res_df[,-1]
  tmp <- apply(gme_res_df,2,as.numeric)
  rownames(tmp) <- rownames(gme_res_df)
  gme_res_df <- as.data.frame(tmp)
  gme_res_df$sig <- -log10(gme_res_df[,4])
  gme_res_df <- gme_res_df[order(-gme_res_df$sig),]
  gme_res_df$`fdr(sc)` <- p.adjust(gme_res_df$`p-value(sc)`)
  out <- list("df"=df,"gme_res_df"=gme_res_df)
  return(out)
}

ttenrich <- function(m,genesets,cores=1) {
  res <- mclapply( 1:length(genesets), function(i) {
    gs <- genesets[i]
    name <- names(gs)
    n_members <- length(which(rownames(m) %in% gs[[1]]))
    if ( n_members > 4 ) {
      tstats <- m[which(rownames(m) %in% gs[[1]]),]
      myn <- length(tstats)
      mymean <- mean(tstats)
      mymedian <- median(tstats)
      wt <- t.test(tstats)
      res <- c(name,myn,mymean,mymedian,wt$p.value)
    }
  } , mc.cores = cores)
  res_df <- do.call(rbind, res)
  rownames(res_df) <- res_df[,1]
  res_df <- res_df[,-1]
  colnames(res_df) <- c("n_genes","t_mean","t_median","p.value")
  tmp <- apply(res_df,2,as.numeric)
  rownames(tmp) <- rownames(res_df)
  res_df <- tmp
  res_df <- as.data.frame(res_df)
  res_df <- res_df[order(res_df$p.value),]
  res_df$logp <- -log10(res_df$p.value )
  res_df$fdr <- p.adjust(res_df$p.value,method="fdr")
  res_df[order(abs(res_df$p.value)),]
  return(res_df)
}

## Load methylation data

@@

design <- as.data.frame(readRDS("dm3_design.Rds"))
dim(design)

mx <- readRDS("dm3_mx.Rds")
dim(mx)

@@

## Split data

Split the data into two smaller groups (s and b).

@@

dess <- design[c(gxs),]
mxs <- mx[,c(gxs)]

desb <- design[c(gxb),]
mxb <- mx[,c(gxb)]

@@

## Limma + GMEA

Using the S dataset, apply limma to conduct differential methylation analysis,
then GMEA.
GMEA has two steps:

1. Aggregate probe t-statistics to genes.

2. Perform enrichment test using 1 sample t-test.

```{r,limmagmea1}

d <- model.matrix(~dess$age+ dess$sex + dess$groups)

fit.reduced <- lmFit(mxs,d)
fit.reduced <- eBayes(fit.reduced)
dm <- topTable(fit.reduced,coef=4, number = Inf)
dma <- merge(myann,dm,by=0)
dma <- dma[order(dma$P.Value),]
dma <- dma[,c("UCSC_RefGene_Name","t")]

dmagg <- agg(dma,cores=8)
m <- dmagg$gme_res_df
m <- m[,"median",drop=FALSE]
tres <- ttenrich(m=m,genesets=gs_symbols,cores=8)
sig <- subset(tres,fdr<0.05)
ssig <- sig

@@

Now repeat with the b set.

```{r,limmagmea2}

d <- model.matrix(~desb$age+ desb$sex + desb$groups)

fit.reduced <- lmFit(mxb,d)
fit.reduced <- eBayes(fit.reduced)
dm <- topTable(fit.reduced,coef=4, number = Inf)
dma <- merge(myann,dm,by=0)
dma <- dma[order(dma$P.Value),]
dma <- dma[,c("UCSC_RefGene_Name","t")]

dmagg <- agg(dma,cores=8)
m <- dmagg$gme_res_df
m <- m[,"median",drop=FALSE]
tres <- ttenrich(m=m,genesets=gs_symbols,cores=8)
sig <- subset(tres,fdr<0.05)
bsig <- sig

@@

ssig_up <- rownames(subset(ssig,t_median>0))
ssig_dn <- rownames(subset(ssig,t_median<0))

bsig_up <- rownames(subset(bsig,t_median>0))
bsig_dn <- rownames(subset(bsig,t_median<0))

v1 <- list("s up"=ssig_up, "s dn"=ssig_dn,
  "b up"=bsig_up,"b dn"=bsig_dn)

@@

* Make sure it is all working properly.

* Repeat for other seed values and aggregate the results.
